# Linear Regression with One Regressor {#lrwor} 

This chapter introduces the basics in linear regression and shows how to perform regression analysis in `r ttcode("R")`. In linear regression, the aim is to model the relationship between a dependent variable $Y$ and one or more explanatory variables denoted by $X_1, X_2, \dots, X_k$. Following the book we will focus on the concept of simple linear regression throughout the whole chapter. In simple linear regression, there is just one explanatory variable $X_1$. <br> 
If, for example, a school cuts its class sizes by hiring new teachers, that is, the school lowers $X_1$, the student-teacher ratios of its classes, how would this affect $Y$, the performance of the students involved in a standardized test? With linear regression we can not only examine whether the student-teacher ratio *does have* an impact on the test results but we can also learn about the *direction* and the *strength* of this effect. 

The following packages are needed for reproducing the code presented in this chapter:

+ `r ttcode("AER")` - accompanies the Book *Applied Econometrics with R* @kleiber2008 and provides useful functions and data sets.

+ `r ttcode("MASS")` - a collection of functions for applied statistics.

Make sure these are installed before you go ahead and try to replicate the examples. The safest way to do so is by checking whether the following code chunk executes without any errors.

```{r, warning=FALSE, message=FALSE, eval=FALSE}
library(AER)
library(MASS)
```

## Simple Linear Regression 

To start with an easy example, consider the following combinations of average test score and the average student-teacher ratio in some fictional school districts.

```{r kable, echo=F, eval=my_output=="html"}
library(knitr)
library(kableExtra)

frame <- data.frame(
  TestScore = c(680, 640, 670, 660, 630, 660, 635),
  STR = c(15, 17, 19, 20, 22, 23.5, 25)
         )
rownames(frame) <- 1:7

t(frame) %>% kable("latex", booktabs = T) %>%
kable_styling(latex_options = "striped")
```

```{r, echo=F, eval=my_output=="latex"}
library(knitr)
library(kableExtra)

frame <- data.frame(
  TestScore = c(680, 640, 670, 660, 630, 660, 635),
  STR = c(15, 17, 19, 20, 22, 23.5, 25)
         )
rownames(frame) <- 1:7

kable(t(frame), "latex", booktabs = T) %>% kable_styling(position = "center")
```

To work with these data in `r ttcode("R")` we begin by generating two vectors: one for the student-teacher ratios (`r ttcode("STR")`) and one for test scores (`r ttcode("TestScore")`), both containing the data from the table above. 

```{r}
# Create sample data
STR <- c(15, 17, 19, 20, 22, 23.5, 25)
TestScore <- c(680, 640, 670, 660, 630, 660, 635) 

# Print out sample data
STR
TestScore
```

In a simple linear regression model, we model the relationship between both variables by a straight line, formally $$ Y = b \cdot X + a. $$ For now, let us suppose that the function which relates test score and student-teacher ratio 
to each other is $$TestScore = 713 - 3 \times STR.$$

It is always a good idea to visualize the data you work with. Here, it is suitable to use `r ttcode("plot()")` to produce a scatterplot with `r ttcode("STR")` on the $x$-axis and `r ttcode("TestScore")` on the $y$-axis. Just call `plot(y_variable ~ x_variable)` whereby `r ttcode("y_variable")` and `r ttcode("x_variable")` are placeholders for the vectors of observations we want to plot. Furthermore, we might want to add a systematic relationship to the plot. To draw a straight line, `r ttcode("R")` provides the function `r ttcode("abline()")`. We just have to call this function with arguments `r ttcode("a")` (representing the intercept) 
and `r ttcode("b")` (representing the slope) after executing `r ttcode("plot()")` in order to add the line to our plot. 

The following code reproduces Figure 4.1 from the textbook. 

<div class="unfolded">
```{r , echo=TRUE, fig.align='center', cache=TRUE}
# create a scatterplot of the data
plot(TestScore ~ STR)

# add the systematic relationship to the plot
abline(a = 713, b = -3)
```
</div>

We find that the line does not touch any of the points although we claimed that it represents the systematic relationship. The reason for this is randomness. Most of the time there are additional influences which imply that there is no bivariate relationship between the two variables. 

In order to account for these differences between observed data and the systematic relationship, we extend our model from above by an *error term* $u$ which captures additional random effects. Put differently, $u$ accounts for all the differences between the regression line and the actual observed data. Beside pure randomness, these deviations could also arise from measurement errors or, as will be discussed later, could be the consequence of leaving out other factors that are relevant in explaining the dependent variable.

Which other factors are plausible in our example? For one thing, the test scores might be driven by the teachers' quality and the background of the students. It is also possible that in some classes, the students were lucky on the test days and thus achieved higher scores. For now, we will summarize such influences by an additive component:

$$ TestScore = \beta_0 + \beta_1 \times STR + \text{other factors} $$

Of course this idea is very general as it can be easily extended to other situations that can be described with a linear model. The basic linear regression function we will work with hence is

$$ Y_i = \beta_0 + \beta_1 X_i + u_i. $$

Key Concept 4.1 summarizes the terminology of the simple linear regression model.   

```{r, eval = my_output == "html", results='asis', echo=FALSE}
cat('
<div class = "keyconcept" id="KC4.1">
<h3 class = "right"> Key Concept 4.1 </h3>          
<h3 class = "left"> Terminology for the Linear Regression Model with a Single Regressor </h3>

<p> The linear regression model is 

$$Y_i = \\beta_0 + \\beta_1 X_1 + u_i$$

where

- the index $i$ runs over the observations, $i=1,\\dots,n$
- $Y_i$ is the *dependent variable*, the *regressand*, or simply the *left-hand variable*
- $X_i$ is the *independent variable*, the *regressor*, or simply the *right-hand variable*
- $Y = \\beta_0 + \\beta_1 X$ is the *population regression line* also called the *population regression function*
- $\\beta_0$ is the *intercept* of the population regression line
- $\\beta_1$ is the *slope* of the population regression line
- $u_i$ is the *error term*.
</p>
</div>
')
```

```{r, eval = my_output == "latex", results='asis', echo=FALSE}
cat('\\begin{keyconcepts}[Terminology for the Linear Regression Model with a Single Regressor]{4.1}
The linear regression model is $$Y_i = \\beta_0 + \\beta_1 X_1 + u_i$$
where
\\begin{itemize}
\\item the index $i$ runs over the observations, $i=1,\\dots,n$
\\item $Y_i$ is the \\textit{dependent variable}, the \\textit{regressand}, or simply the \\textit{left-hand variable}
\\item $X_i$ is the \\textit{independent variable}, the \\textit{regressor}, or simply the \\textit{right-hand variable}
\\item $Y = \\beta_0 + \\beta_1 X$ is the \\textit{population regression line} also called the \\textit{population regression function}
\\item $\\beta_0$ is the \\textit{intercept} of the population regression line
\\item $\\beta_1$ is the \\textit{slope} of the population regression line
\\item $u_i$ is the \\textit{error term}.
\\end{itemize}
\\end{keyconcepts}
')
```

## Estimating the Coefficients of the Linear Regression Model

In practice, the intercept $\beta_0$ and slope $\beta_1$ of the population regression line are unknown. Therefore, we must employ data to estimate both unknown parameters. In the following, a real world example will be used to demonstrate how this is achieved. We want to relate test scores to student-teacher ratios measured in Californian schools. The test score is the district-wide average of reading and math scores for fifth graders. Again, the class size is measured as the number of students divided by the number of teachers (the student-teacher ratio). As for the data, the California School data set (`r ttcode("CASchools")`) comes with an `r ttcode("R")` package called `r ttcode("AER")`, an acronym for [Applied Econometrics with R](https://cran.r-project.org/web/packages/AER/AER.pdf) [@R-AER]. After installing the package with `r ttcode('install.packages("AER")')` and attaching it with `r ttcode('library(AER)')` the data set can be loaded using the function `r ttcode("data()")`. 

```{r, message=FALSE, warning=FALSE, eval=FALSE}
# install the AER package (once)
install.packages("AER")

# load the AER package 
library(AER)   

# load the the data set in the workspace
data(CASchools) 
```

```{r, message=FALSE, warning=FALSE,echo=FALSE}
# load the AER package 
library(AER)   

# load the the data set in the workspace
data(CASchools) 
```

Once a package has been installed it is available for use at further occasions when invoked with `r ttcode("library()")` --- there is no need to run `r ttcode("install.packages()")` again!

It is interesting to know what kind of object we are dealing with. 
`r ttcode("class()")` returns the class of an object. Depending on the class of an object some functions (for example `r ttcode("plot()")` and `r ttcode("summary()")`) behave differently.

Let us check the class of the object `r ttcode("CASchools")`.

```{r}
class(CASchools)
```

It turns out that `r ttcode("CASchools")` is of class `r ttcode("data.frame")` which is a convenient format to work with, especially for performing regression analysis. 

With help of `r ttcode("head()")` we get a first overview of our data. This function shows only the first 6 rows of the data set which prevents an overcrowded console output. 

```{block2, note-text, type='rmdnote'}
Press <tt>ctrl + L</tt> to clear the console. This command deletes any code that has been typed in and executed by you or printed to the console by <tt>R</tt> functions. The good news is that anything else is left untouched. You neither loose defined variables etc. nor the code history. It is still possible to recall previously executed <tt>R</tt> commands using the up and down keys. If you are working in *RStudio*, press <tt>ctrl + Up</tt> on your keyboard (<tt>CMD + Up</tt> on a Mac) to review a list of previously entered commands.
```

```{r}
head(CASchools)
```

We find that the data set consists of plenty of variables and that most of them are numeric. 

By the way: an alternative to `r ttcode("class()")` and `r ttcode("head()")` is `r ttcode("str()")` which is deduced from 'structure' and gives a comprehensive overview of the object. Try!

Turning back to `r ttcode("CASchools")`, the two variables we are interested in (i.e., average test score and the student-teacher ratio) are *not* included. However, it is possible to calculate both from the provided data. To obtain the student-teacher ratios, we simply divide the number of students by the number of teachers. The average test score is the arithmetic mean of the test score for reading and the score of the math test. The next code chunk shows how the two variables can be constructed as vectors and how they are appended to `r ttcode("CASchools")`.

```{r}
# compute STR and append it to CASchools
CASchools$STR <- CASchools$students/CASchools$teachers 

# compute TestScore and append it to CASchools
CASchools$score <- (CASchools$read + CASchools$math)/2     
```

If we ran `r ttcode("head(CASchools)")` again we would find the two variables of interest as additional columns named `r ttcode("STR")` and `r ttcode("score")` (check this!).

Table 4.1 from the textbook summarizes the distribution of test scores and student-teacher ratios. There are several functions which can be used to produce similar results, e.g.,

- `r ttcode("mean()")` (computes the arithmetic mean of the provided numbers),

- `r ttcode("sd()")` (computes the sample standard deviation),

- `r ttcode("quantile()")` (returns a vector of the specified sample quantiles for the data). 

The next code chunk shows how to achieve this. First, we compute summary statistics on the columns `r ttcode("STR")` and `r ttcode("score")` of `r ttcode("CASchools")`. In order to get nice output we gather the measures in a `r ttcode("data.frame")` named `r ttcode("DistributionSummary")`. 

```{r Table 4.1, results='hold'}
# compute sample averages of STR and score
avg_STR <- mean(CASchools$STR) 
avg_score <- mean(CASchools$score)

# compute sample standard deviations of STR and score
sd_STR <- sd(CASchools$STR) 
sd_score <- sd(CASchools$score)

# set up a vector of percentiles and compute the quantiles 
quantiles <- c(0.10, 0.25, 0.4, 0.5, 0.6, 0.75, 0.9)
quant_STR <- quantile(CASchools$STR, quantiles)
quant_score <- quantile(CASchools$score, quantiles)

# gather everything in a data.frame 
DistributionSummary <- data.frame(Average = c(avg_STR, avg_score), 
                                  StandardDeviation = c(sd_STR, sd_score), 
                                  quantile = rbind(quant_STR, quant_score))

# print the summary to the console
DistributionSummary
```

As for the sample data, we use `r ttcode("plot()")`. This allows us to detect characteristics of our data, such as outliers which are harder to discover by looking at mere numbers. This time we add some additional arguments to the call of `r ttcode("plot()")`.

The first argument in our call of `r ttcode("plot()")`, `r ttcode("score ~ STR")`, is again a formula that states variables on the y- and the x-axis. However, this time the two variables are not saved in separate vectors but are columns of `r ttcode("CASchools")`. Therefore, `r ttcode("R")` would not find them without the argument `r ttcode("data")` being correctly specified. `r ttcode("data")` must be in accordance with the name of the `r ttcode("data.frame")` to which the variables belong to, in this case `r ttcode("CASchools")`. Further arguments are used to change the appearance of the plot: while `r ttcode("main")` adds a title, `r ttcode("xlab")` and `r ttcode("ylab")` add custom labels to both axes.  

<div class="unfolded">
```{r, fig.align='center'}
plot(score ~ STR, 
     data = CASchools,
     main = "Scatterplot of TestScore and STR", 
     xlab = "STR (X)",
     ylab = "Test Score (Y)")
```
</div>

The plot (Figure 4.2 in the book) shows the scatterplot of all observations on the student-teacher ratio and test score. We see that the points are strongly scattered, and that the variables are negatively correlated. That is, we expect to observe lower test scores in bigger classes.

The function `r ttcode("cor()")` (see `r ttcode("?cor")` for further info) can be used to compute the correlation between two *numeric* vectors. 

```{r}
cor(CASchools$STR, CASchools$score)
```

As the scatterplot already suggests, the correlation is negative but rather weak. 

The task we are now facing is to find a line which best fits the data. Of course we could simply stick with graphical inspection and correlation analysis and then select the best fitting line by eyeballing. However, this would be rather subjective: different observers would draw different regression lines. On this account, we are interested in techniques that are less arbitrary. Such a technique is given by ordinary least squares (OLS) estimation.   

### The Ordinary Least Squares Estimator {-}

The OLS estimator chooses the regression coefficients such that the estimated regression line is as "close" as possible to the observed data points. Here, closeness is measured by the sum of the squared mistakes made in predicting $Y$ given $X$. Let $b_0$ and $b_1$ be some estimators of $\beta_0$ and $\beta_1$. Then the sum of squared estimation mistakes can be expressed as 

$$ \sum^n_{i = 1} (Y_i - b_0 - b_1 X_i)^2. $$

The OLS estimator in the simple regression model is the pair of estimators for intercept and slope which minimizes the expression above. The derivation of the OLS estimators for both parameters are presented in Appendix 4.1 of the book. The results are summarized in Key Concept 4.2.

```{r, eval = my_output == "html", results='asis', echo=FALSE}
cat('
<div class = "keyconcept" id="KC4.2">
<h3 class = "right"> Key Concept 4.2 </h3>          
<h3 class = "left"> The OLS Estimator, Predicted Values, and Residuals </h3>
<p> The OLS estimators of the slope $\\beta_1$ and the intercept $\\beta_0$ in the simple linear regression model are
\\begin{align}
  \\hat\\beta_1 & = \\frac{ \\sum_{i = 1}^n (X_i - \\overline{X})(Y_i - \\overline{Y}) } { \\sum_{i=1}^n (X_i - \\overline{X})^2},  \\\\
  \\\\
  \\hat\\beta_0 & =  \\overline{Y} - \\hat\\beta_1 \\overline{X}. 
\\end{align}
The OLS predicted values $\\widehat{Y}_i$ and residuals $\\hat{u}_i$ are
\\begin{align}
  \\widehat{Y}_i & =  \\hat\\beta_0 + \\hat\\beta_1 X_i,\\\\
  \\\\
  \\hat{u}_i & =  Y_i - \\widehat{Y}_i. 
\\end{align}

The estimated intercept $\\hat{\\beta}_0$, the slope parameter $\\hat{\\beta}_1$ and the residuals $\\left(\\hat{u}_i\\right)$ are computed from a sample of $n$ observations of $X_i$ and $Y_i$, $i$, $...$,  $n$. These are *estimates* of the unknown population intercept $\\left(\\beta_0 \\right)$, slope $\\left(\\beta_1\\right)$, and error term $(u_i)$.
</p>

The formulas presented above may not be very intuitive at first glance. The following interactive application aims to help you understand the mechanics of OLS. You can add observations by clicking into the coordinate system where the data are represented by points. Once two or more observations are available, the application computes a regression line using OLS and some statistics which are displayed in the right panel. The results are updated as you add further observations to the left panel. A double-click resets the application, i.e., all data are removed.

<iframe height="410" width="900" frameborder="0" scrolling="no" src="SimpleRegression.html"></iframe>

</div>')
```

```{r, eval = my_output == "latex", results='asis', echo=FALSE}
cat('\\begin{keyconcepts}[The OLS Estimator, Predicted Values, and Residuals]{4.2}
The OLS estimators of the slope $\\beta_1$ and the intercept $\\beta_0$ in the simple linear regression model are
\\begin{align*}
  \\hat\\beta_1 & = \\frac{ \\sum_{i = 1}^n (X_i - \\overline{X})(Y_i - \\overline{Y}) } { \\sum_{i=1}^n (X_i - \\overline{X})^2},  \\\\
\\hat\\beta_0 & =  \\overline{Y} - \\hat\\beta_1 \\overline{X}. 
\\intertext{The OLS predicted values $\\widehat{Y}_i$ and residuals $\\hat{u}_i$ are}
  \\widehat{Y}_i & =  \\hat\\beta_0 + \\hat\\beta_1 X_i,\\\\
  \\hat{u}_i & =  Y_i - \\widehat{Y}_i. 
\\end{align*}

The estimated intercept $\\hat{\\beta}_0$, the slope parameter $\\hat{\\beta}_1$ and the residuals $\\left(\\hat{u}_i\\right)$ are computed from a sample of $n$ observations of $X_i$ and $Y_i$, $i$, $...$,  $n$. These are \\textit{estimates} of the unknown true population intercept $\\left(\\beta_0 \\right)$, slope $\\left(\\beta_1\\right)$, and error term $(u_i)$.
\\end{keyconcepts}
')
```

There are many possible ways to compute $\hat{\beta}_0$ and $\hat{\beta}_1$ in `r ttcode("R")`. For example, we could implement the formulas presented in Key Concept 4.2 with two of `r ttcode("R")`'s most basic functions: `r ttcode("mean()")` and `r ttcode("sum()")`.

<div class="unfolded">
```{r, echo=-1}
rm(STR)
attach(CASchools) # allows to use the variables contained in CASchools directly

# compute beta_1_hat
beta_1 <- sum((STR - mean(STR)) * (score - mean(score))) / sum((STR - mean(STR))^2)

# compute beta_0_hat
beta_0 <- mean(score) - beta_1 * mean(STR)

# print the results to the console
beta_1
beta_0
```
</div>

Of course, there are even more manual ways to perform these tasks. With OLS being one of the most widely-used estimation techniques, `r ttcode("R")` of course already contains a built-in function named `r ttcode("lm()")` (**l**inear **m**odel) which can be used to carry out regression analysis.

The first argument of the function to be specified is, similar to `r ttcode("plot()")`, the regression formula with the basic syntax `r ttcode("y ~ x")` where `r ttcode("y")` is the dependent variable and `r ttcode("x")` the explanatory variable. The argument `r ttcode("data")` determines the data set to be used in the regression. We now revisit the example from the book where the relationship between the test scores and the class sizes is analyzed. The following code uses `r ttcode("lm()")` to replicate the results presented in figure 4.3 of the book. 

```{r}
# estimate the model and assign the result to linear_model
linear_model <- lm(score ~ STR, data = CASchools)

# print the standard output of the estimated lm object to the console 
linear_model
```

Let us add the estimated regression line to the plot. This time we also enlarge the ranges of both axes by setting the arguments `r ttcode("xlim")` and `r ttcode("ylim")`.

<div class="unfolded">
```{r, fig.align='center'}
# plot the data
plot(score ~ STR, 
     data = CASchools,
     main = "Scatterplot of TestScore and STR", 
     xlab = "STR (X)",
     ylab = "Test Score (Y)",
     xlim = c(10, 30),
     ylim = c(600, 720))

# add the regression line
abline(linear_model) 
```
</div>

Did you notice that this time, we did not pass the intercept and slope parameters to `r ttcode("abline")`? If you call `r ttcode("abline()")` on an object of class `r ttcode("lm")` that only contains a single regressor, `r ttcode("R")` draws the regression line automatically! 

## Measures of Fit

After fitting a linear regression model, a natural question is how well the model describes the data. Visually, this amounts to assessing whether the observations are tightly clustered around the regression line. Both the *coefficient of determination* and the *standard error of the regression* measure how well the OLS Regression line fits the data. 

### The Coefficient of Determination {-}

$R^2$, the *coefficient of determination*, is the fraction of the sample variance of $Y_i$ that is explained by $X_i$. Mathematically, the $R^2$ can be written as the ratio of the explained sum of squares to the total sum of squares. The *explained sum of squares* ($ESS$) is the sum of squared deviations of the predicted values $\hat{Y_i}$, from the average of the $Y_i$. The *total sum of squares* ($TSS$) is the sum of squared deviations of the $Y_i$ from their average. Thus we have 

\begin{align}
  ESS & =  \sum_{i = 1}^n \left( \hat{Y_i} - \overline{Y} \right)^2,   \\
  TSS & =  \sum_{i = 1}^n \left( Y_i - \overline{Y} \right)^2,   \\
  R^2 & = \frac{ESS}{TSS}.
\end{align}

Since $TSS = ESS + SSR$ we can also write

$$ R^2 = 1- \frac{SSR}{TSS} $$ 

where $SSR$ is the sum of squared residuals, a measure for the errors made when predicting the $Y$ by $X$. The $SSR$ is defined as

$$ SSR = \sum_{i=1}^n \hat{u}_i^2. $$

$R^2$ lies between $0$ and $1$. It is easy to see that a perfect fit, i.e., no errors made when fitting the regression line, implies $R^2 = 1$ since then we have $SSR=0$. On the contrary, if our estimated regression line does not explain any variation in the $Y_i$, we have $ESS=0$ and consequently $R^2=0$.  

### The Standard Error of the Regression {-}

The *Standard Error of the Regression* ($SER$) is an estimator of the standard deviation of the residuals $\hat{u}_i$. As such it measures the magnitude of a typical deviation from the regression line, i.e., the magnitude of a typical residual.

$$ SER = s_{\hat{u}} = \sqrt{s_{\hat{u}}^2} \ \ \ \text{where} \ \ \ s_{\hat{u} }^2 = \frac{1}{n-2} \sum_{i = 1}^n \hat{u}^2_i = \frac{SSR}{n - 2} $$

Remember that the $u_i$ are *unobserved*. This is why we use their estimated counterparts, the residuals $\hat{u}_i$, instead. See Chapter 4.3 of the book for a more detailed comment on the $SER$.

### Application to the Test Score Data {-}

Both measures of fit can be obtained by using the  function `r ttcode("summary()")` with an `r ttcode("lm")` object provided as the only argument. While the function `r ttcode("lm()")` only prints out the estimated coefficients to the console, `r ttcode("summary()")` provides additional predefined information such as the regression's $R^2$ and the $SER$.  

<div class="unfolded">
```{r}
mod_summary <- summary(linear_model)
mod_summary
```
</div>

The $R^2$ in the output is called *Multiple R-squared* and has a value of $0.051$. Hence, $5.1 \%$ of the variance of the dependent variable $score$ is explained by the explanatory variable $STR$. That is, the regression explains little of the variance in $score$, and much of the variation in test scores remains unexplained (cf. Figure 4.3 of the book).   

The $SER$ is called *Residual standard error* and equals $18.58$. The unit of the $SER$ is the same as the unit of the dependent variable. That is, on average the deviation of the actual achieved test score and the regression line is $18.58$ points. 

Now, let us check whether `r ttcode("summary()")` uses the same definitions for $R^2$ and $SER$ as we do when computing them manually.

```{r}
# compute R^2 manually
SSR <- sum(mod_summary$residuals^2)
TSS <- sum((score - mean(score))^2)
R2 <- 1 - SSR/TSS

# print the value to the console
R2

# compute SER manually
n <- nrow(CASchools)
SER <- sqrt(SSR / (n-2))

# print the value to the console
SER
```

We find that the results coincide. Note that the values provided by `r ttcode("summary()")` are rounded to two decimal places. Can you do so using `r ttcode("R")`?

## The Least Squares Assumptions {#tlsa}

OLS performs well under a quite broad variety of different circumstances. However, there are some assumptions which need to be satisfied in order to ensure that the estimates are normally distributed in large samples (we discuss this in Chapter \@ref{tsdotoe}).

```{r, eval = my_output == "html", results='asis', echo=FALSE}
cat('
<div class = "keyconcept" id="KC4.3">
<h3 class = "right"> Key Concept 4.3 </h3>          
<h3 class = "left"> The Least Squares Assumptions </h3>
<p> 
$$Y_i = \\beta_0 + \\beta_1 X_i + u_i \\text{, } i = 1,\\dots,n$$
where

1. The error term $u_i$ has conditional mean zero given $X_i$: $E(u_i|X_i) = 0$.
2. $(X_i,Y_i), i = 1,\\dots,n$ are independent and identically distributed (i.i.d.) draws from their joint distribution.
3. Large outliers are unlikely: $X_i$ and $Y_i$ have nonzero finite fourth moments.
</p>
</div>
')
```

```{r, eval = my_output == "latex", results='asis', echo=FALSE}
cat('\\begin{keyconcepts}[The Least Squares Assumptions]{4.3}
$$Y_i = \\beta_0 + \\beta_1 X_i + u_i \\text{, } i = 1,\\dots,n$$
where

\\begin{enumerate}
\\item The error term $u_i$ has conditional mean zero given $X_i$: $E(u_i|X_i) = 0$.
\\item $(X_i,Y_i), i = 1,\\dots,n$ are independent and identically distributed (i.i.d.) draws from their joint distribution.
\\item Large outliers are unlikely: $X_i$ and $Y_i$ have nonzero finite fourth moments.
\\end{enumerate}
\\end{keyconcepts}
')
```

### Assumption 1: The Error Term has Conditional Mean of Zero {-}

This means that no matter which value we choose for $X$, the error term $u$ must not show any systematic pattern and must have a mean of $0$.
Consider the case that, unconditionally, $E(u) = 0$, but for low and high values of $X$, the error term tends to be positive and for midrange values of
$X$ the error tends to be negative. We can use R to construct such an example. To do so we generate our own data using `r ttcode("R")`'s built-in random number generators. 

We will use the following functions:

* `r ttcode("runif()")` - generates uniformly distributed random numbers
* `r ttcode("rnorm()")` - generates normally distributed random numbers
* `r ttcode("predict()")` - does predictions based on the results of model fitting functions like `r ttcode("lm()")`
* `r ttcode("lines()")` - adds line segments to an existing plot

We start by creating a vector containing values that are uniformly distributed on the interval $[-5,5]$. This can be done with the function `r ttcode("runif()")`. We also need to simulate the error term. For this we generate normally distributed random numbers with a mean equal to $0$ and a variance of $1$ using `r ttcode("rnorm()")`. The $Y$ values are obtained as a quadratic function of the $X$ values and the error.

After generating the data we estimate both a simple regression model and a quadratic model that also includes the regressor $X^2$ (this is a multiple regression model, see Chapter \@ref(rmwmr)). Finally, we plot the simulated data and add a the estimated regression line of a simple regression model as well as the predictions made with a quadratic model to compare the fit graphically.

<div class="unfolded">
```{r, fig.align='center'}
# set a seed to make the results reproducible
set.seed(321)

# simulate the data 
X <- runif(50, min = -5, max = 5)
u <- rnorm(50, sd = 5)  

# the true relation  
Y <- X^2 + 2 * X + u                

# estimate a simple regression model 
mod_simple <- lm(Y ~ X)

# predict using a quadratic model 
prediction <- predict(lm(Y ~ X + I(X^2)), data.frame(X = sort(X)))

# plot the results
plot(Y ~ X)
abline(mod_simple, col = "red")
lines(sort(X), prediction)
```
</div>

The plot shows what is meant by $E(u_i|X_i) = 0$ and why it does not hold for the linear model: 

Using the quadratic model (represented by the black curve) we see that there are no systematic deviations of the observation from the predicted relation. It is credible that the assumption is not violated when such a model is employed. However, using a simple linear regression model we see that the assumption is probably violated as $E(u_i|X_i)$ varies with the $X_i$.

### Assumption 2: Independently and Identically Distributed Data {-}

Most sampling schemes used when collecting data from populations produce i.i.d.-samples. For example, we could use `r ttcode("R")`'s random number generator to randomly select student IDs from a university's enrollment list and record age $X$ and earnings $Y$ of the corresponding students. This is a typical example of simple random sampling and ensures that all the $(X_i, Y_i)$ are drawn randomly from the same population.

A prominent example where the i.i.d. assumption is not fulfilled is time series data where we have observations on the same unit over time. For example, take $X$ as the number of workers in a production company over time. Due to business transformations, the company cuts job periodically by a specific share but there are also some non-deterministic influences that relate to economics, politics etc. Using `r ttcode("R")` we can easily simulate such a process and plot it. 

We start the series with a total of 5000 workers and simulate the reduction of employment with an autoregressive process that exhibits a downward movement in the long-run and has normally distributed errors:^[See Chapter \@ref(ittsraf) for more on autoregressive processes and time series analysis in general.]

$$ employment_t = -5 + 0.98 \cdot employment_{t-1} + u_t $$
<div class="unfolded"> 
```{r, fig.align="center"}
# set seed
set.seed(123)

# generate a date vector
Date <- seq(as.Date("1951/1/1"), as.Date("2000/1/1"), "years")

# initialize the employment vector
X <- c(5000, rep(NA, length(Date)-1))

# generate time series observations with random influences
for (i in 2:length(Date)) {
  
    X[i] <- -50 + 0.98 * X[i-1] + rnorm(n = 1, sd = 200)
    
}

#plot the results
plot(x = Date, 
     y = X, 
     type = "l", 
     col = "steelblue", 
     ylab = "Workers", 
     xlab = "Time")
```
</div>

It is evident that the observations on the number of employees cannot be independent in this example: the level of today's employment is correlated with tomorrows employment level. Thus, the i.i.d. assumption is violated.

### Assumption 3: Large Outliers are Unlikely {-}

It is easy to come up with situations where extreme observations, i.e., observations that deviate considerably from the usual range of the data, may occur. Such observations are called outliers. Technically speaking, assumption 3 requires that $X$ and $Y$ have a finite kurtosis.^[See Chapter 4.4 of the book.]

Common cases where we want to exclude or (if possible) correct such outliers is when they are apparently typos, conversion errors or measurement errors. Even if it seems like extreme observations have been recorded correctly, it is advisable to exclude them before estimating a model since OLS suffers from *sensitivity to outliers*.

What does this mean? One can show that extreme observations receive heavy weighting in the estimation of the unknown regression coefficients when using OLS. Therefore, outliers can lead to strongly distorted estimates of regression coefficients. To get a better impression of this issue, consider the following application where we have placed some sample data on $X$ and $Y$ which are highly correlated. The relation between $X$ and $Y$ seems to be explained pretty good by the plotted regression line: all of the blue dots lie close to the red line and we have $R^2=0.92$.

Now go ahead and add a further observation at, say, $(18,2)$. This observations clearly is an outlier. The result is quite striking: the estimated regression line differs greatly from the one we adjudged to fit the data well. The slope is heavily downward biased and $R^2$ decreased to a mere $29\%$! <br>
Double-click inside the coordinate system to reset the app. Feel free to experiment. Choose different coordinates for the outlier or add additional ones.

<iframe height="410" width="900" frameborder="0" scrolling="no" src="Outlier.html"></iframe>

The following code roughly reproduces what is shown in figure 4.5 in the book. As done above we use sample data generated using `r ttcode("R")`'s random number functions `r ttcode("rnorm()")` and `r ttcode("runif()")`. We estimate two simple regression models, one based on the original data set and another using a modified set where one observation is change to be an outlier and then plot the results. In order to understand the complete code you should be familiar with the function `r ttcode("sort()")` which sorts the entries of a numeric vector in ascending order.

<div class="unfolded">
```{r, fig.align='center'}
# set seed
set.seed(123)

# generate the data
X <- sort(runif(10, min = 30, max = 70))
Y <- rnorm(10 , mean = 200, sd = 50)
Y[9] <- 2000

# fit model with outlier
fit <- lm(Y ~ X)

# fit model without outlier
fitWithoutOutlier <- lm(Y[-9] ~ X[-9])

# plot the results
plot(Y ~ X)
abline(fit)
abline(fitWithoutOutlier, col = "red")
```
</div>

## The Sampling Distribution of the OLS Estimator {#tsdotoe} 

Because $\hat{\beta}_0$ and $\hat{\beta}_1$ are computed from a sample, the estimators themselves are random variables with a probability distribution --- the so-called sampling distribution of the estimators --- which describes the values they could take on over different samples. Although the sampling distribution of $\hat\beta_0$ and $\hat\beta_1$ can be complicated when the sample size is small and generally changes with the number of observations, $n$, it is possible, provided the assumptions discussed in the book are valid, to make certain statements about it that hold for all $n$. In particular 
$$ E(\hat{\beta}_0) = \beta_0 \ \ \text{and} \ \  E(\hat{\beta}_1) = \beta_1,$$
that is, $\hat\beta_0$ and $\hat\beta_1$ are unbiased estimators of $\beta_0$ and $\beta_1$, the true parameters. If the sample is sufficiently large, by the central limit theorem the *joint* sampling distribution of the estimators is well approximated by the bivariate normal distribution (<a href="#mjx-eqn-2.1">2.1</a>). This implies that the marginal distributions are also normal in large samples. Core facts on the large-sample distributions of $\hat\beta_0$ and $\hat\beta_1$ are presented in Key Concept 4.4. 

```{r, eval = my_output == "html", results='asis', echo=FALSE}
cat('
<div class = "keyconcept" id="KC4.4">
<h3 class = "right"> Key Concept 4.4 </h3>          
<h3 class = "left"> Large Sample Distribution of $\\hat\\beta_0$ and $\\hat\\beta_1$ </h3>

<p> If the least squares assumptions in Key Concept 4.3 hold, then in large samples $\\hat\\beta_0$ and $\\hat\\beta_1$ have a joint normal sampling distribution. The large sample normal distribution of $\\hat\\beta_1$ is $\\mathcal{N}(\\beta_1, \\sigma^2_{\\hat\\beta_1})$, where the variance of the distribution, $\\sigma^2_{\\hat\\beta_1}$, is 

\\begin{align}
\\sigma^2_{\\hat\\beta_1} = \\frac{1}{n} \\frac{Var \\left[ \\left(X_i - \\mu_X \\right) u_i  \\right]}  {\\left[  Var \\left(X_i \\right)  \\right]^2}. (\\#eq:olsvar1)
\\end{align}

The large sample normal distribution of $\\hat\\beta_0$ is $\\mathcal{N}(\\beta_0, \\sigma^2_{\\hat\\beta_0})$ with

\\begin{align}
\\sigma^2_{\\hat\\beta_0} =  \\frac{1}{n} \\frac{Var \\left( H_i u_i \\right)}{ \\left[  E \\left(H_i^2  \\right)  \\right]^2 } \\ , \\ \\text{where} \\ \\ H_i = 1 - \\left[ \\frac{\\mu_X} {E \\left( X_i^2\\right)} \\right] X_i. (\\#eq:olsvar2)
\\end{align}

The interactive simulation below continuously generates random samples $(X_i,Y_i)$ of $200$ observations where $E(Y\\vert X) = 100 + 3X$, estimates a simple regression model, stores the estimate of the slope $\\beta_1$ and visualizes the distribution of the $\\widehat{\\beta}_1$s observed so far using a histogram. The idea here is that for a large number of $\\widehat{\\beta}_1$s, the histogram gives a good approximation of the sampling distribution of the estimator. By decreasing the time between two sampling iterations, it becomes clear that the shape of the histogram approaches the characteristic bell shape of a normal distribution centered at the true slope of $3$.

<iframe height="470" width="700" frameborder="0" scrolling="no" src="SmallSampleDIstReg.html"></iframe>

</p>
</div>
')
```

```{r, eval = my_output == "latex", results='asis', echo=FALSE}
cat('\\begin{keyconcepts}[Large Sample Distribution of $\\hat\\beta_0$ and $\\hat\\beta_1$]{4.4}
If the least squares assumptions in Key Concept 4.3 hold, then in large samples $\\hat\\beta_0$ and $\\hat\\beta_1$ have a joint normal sampling distribution. The large sample normal distribution of $\\hat\\beta_1$ is $\\mathcal{N}(\\beta_1, \\sigma^2_{\\hat\\beta_1})$, where the variance of the distribution, $\\sigma^2_{\\hat\\beta_1}$, is 

\\begin{equation}
\\sigma^2_{\\hat\\beta_1} = \\frac{1}{n} \\frac{Var \\left[ \\left(X_i - \\mu_X \\right) u_i  \\right]}  {\\left[  Var \\left(X_i \\right)  \\right]^2}.
\\end{equation}

The large sample normal distribution of $\\hat\\beta_0$ is $\\mathcal{N}(\\beta_0, \\sigma^2_{\\hat\\beta_0})$ with

\\begin{equation}
\\sigma^2_{\\hat\\beta_0} =  \\frac{1}{n} \\frac{Var \\left( H_i u_i \\right)}{ \\left[  E \\left(H_i^2  \\right)  \\right]^2 } \\ , \\ \\text{where} \\ \\ H_i = 1 - \\left[ \\frac{\\mu_X} {E \\left( X_i^2\\right)} \\right] X_i.
\\end{equation}

The interactive simulation below continuously generates random samples $(X_i,Y_i)$ of $200$ observations where $E(Y\\vert X) = 100 + 3X$, estimates a simple regression model, stores the estimate of the slope $\\beta_1$ and visualizes the distribution of the $\\widehat{\\beta}_1$s observed so far using a histogram. The idea here is that for a large number of $\\widehat{\\beta}_1$s, the histogram gives a good approximation of the sampling distribution of the estimator. By decreasing the time between two sampling iterations, it becomes clear that the shape of the histogram approaches the characteristic bell shape of a normal distribution centered at the true slope of $3$.\\vspace{0.5cm}

\\begin{center}\\textit{This interactive part of the book is only available in the HTML version.}\\end{center}

\\end{keyconcepts}
')
```

### Simulation Study 1 {-}

Whether the statements of Key Concept 4.4 really hold can also be verified using `r ttcode("R")`. For this we first we build our own population of $100000$ observations in total. To do this we need values for the independent variable $X$, for the error term $u$, and for the parameters $\beta_0$ and $\beta_1$. With these combined in a simple regression model, we compute the dependent variable $Y$. <br> In our example we generate the numbers $X_i$, $i = 1$, ... ,$100000$ by drawing a random sample from a uniform distribution on the interval $[0,20]$. The realizations of the error terms $u_i$ are drawn from a standard normal distribution with parameters $\mu = 0$ and $\sigma^2 = 100$ (note that `r ttcode("rnorm()")` requires $\sigma$ as input for the argument `r ttcode("sd")`, see `r ttcode("?rnorm")`). Furthermore we chose $\beta_0 = -2$ and $\beta_1 = 3.5$ so the true model is

$$ Y_i = -2 + 3.5 \cdot X_i. $$

Finally, we store the results in a data.frame.  

```{r}
# simulate data
N <- 100000
X <- runif(N, min = 0, max = 20)
u <- rnorm(N, sd = 10)

# population regression
Y <- -2 + 3.5 * X + u
population <- data.frame(X, Y)
```

From now on we will consider the previously generated data as the true population (which of course would be *unknown* in a real world application, otherwise there would be no reason to draw a random sample in the first place). The knowledge about the true population and the true relationship between $Y$ and $X$ can be used to verify the statements made in Key Concept 4.4. 

First, let us calculate the true variances $\sigma^2_{\hat{\beta}_0}$ and $\sigma^2_{\hat{\beta}_1}$ for a randomly drawn sample of size $n = 100$. 

```{r, fig.align='center'}
# set sample size
n <- 100

# compute the variance of beta_hat_0
H_i <- 1 - mean(X) / mean(X^2) * X
var_b0 <- var(H_i * u) / (n * mean(H_i^2)^2 )

# compute the variance of hat_beta_1
var_b1 <- var( ( X - mean(X) ) * u ) / (100 * var(X)^2)
```

```{r}
# print variances to the console
var_b0
var_b1
```

Now let us assume that we do not know the true values of $\beta_0$ and $\beta_1$ and that it is not possible to observe the whole population. However, we can observe a random sample of $n$ observations. Then, it would not be possible to compute the true parameters but we could obtain estimates of $\beta_0$ and $\beta_1$ from the sample data using OLS. However, we know that these estimates are outcomes of random variables themselves since the observations are randomly sampled from the population. Key Concept 4.4 describes their distributions for large $n$. When drawing a single sample of size $n$ it is not possible to make any statement about these distributions. Things change if we repeat the sampling scheme many times and compute the estimates for each sample: using this procedure we simulate outcomes of the respective distributions.    

To achieve this in R, we employ the following approach:

- We assign the number of repetitions, say $10000$, to `r ttcode("reps")` and then initialize a matrix `r ttcode("fit")` were the estimates obtained in each sampling iteration shall be stored row-wise. Thus `r ttcode("fit")` has to be a matrix of dimensions `r ttcode("reps")`$\times2$.
- In the next step we draw `r ttcode("reps")` random samples of size `r ttcode("n")` from the population and obtain the OLS estimates for each sample. The results are stored as row entries in the outcome matrix `r ttcode("fit")`. This is done using a `r ttcode("for()")` loop.
- At last, we estimate variances of both estimators using the sampled outcomes and plot histograms of the latter. We also add a plot of the density functions belonging to the distributions that follow from Key Concept 4.4. The function `r ttcode("bquote()")` is used to obtain math expressions in the titles and labels of both plots. See `r ttcode("?bquote")`.

```{r, cache=T}
# set repetitions and sample size
n <- 100
reps <- 10000

# initialize the matrix of outcomes
fit <- matrix(ncol = 2, nrow = reps)

# loop sampling and estimation of the coefficients
for (i in 1:reps){
  
 sample <- population[sample(1:N, n), ]
 fit[i, ] <- lm(Y ~ X, data = sample)$coefficients
 
}

# compute variance estimates using outcomes
var(fit[, 1])
var(fit[, 2])
```

<div class="unfolded"> 
```{r, fig.align='center'}
# divide plotting area as 1-by-2 array
par(mfrow = c(1, 2))

# plot histograms of beta_0 estimates
hist(fit[, 1],
     cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ 10000 ~ beta[0] ~ Estimates), 
     xlab = bquote(hat(beta)[0]), 
     freq = F)

# add true distribution to plot
curve(dnorm(x, 
            -2, 
            sqrt(var_b0)), 
      add = T, 
      col = "darkred")

# plot histograms of beta_hat_1 
hist(fit[, 2],
    cex.main = 1,
     main = bquote(The ~ Distribution  ~ of ~ 10000 ~ beta[1] ~ Estimates), 
     xlab = bquote(hat(beta)[1]), 
     freq = F)

# add true distribution to plot
curve(dnorm(x, 
            3.5, 
            sqrt(var_b1)), 
      add = T, 
      col = "darkred")
```
</div>

Our variance estimates support the statements made in Key Concept 4.4, coming close to the theoretical values. The histograms suggest that the distributions of the estimators can be well approximated by the respective theoretical normal distributions stated in Key Concept 4.4.    

### Simulation Study 2 {-}

A further result implied by Key Concept 4.4 is that both estimators are consistent, i.e., they converge in probability to the true parameters we are interested in. This is because they are asymptotically unbiased and their variances converge to $0$ as $n$ increases. We can check this by repeating the simulation above for a sequence of increasing sample sizes. This means we no longer assign the sample size but a *vector* of sample sizes: `r ttcode("n <- c(...)")`. <br>
Let us look at the distributions of $\beta_1$. The idea here is to add an additional call of `r ttcode("for()")` to the code. This is done in order to loop over the vector of sample sizes `r ttcode("n")`. For each of the sample sizes we carry out the same simulation as before but plot a density estimate for the outcomes of each iteration over `r ttcode("n")`. Notice that we have to change `r ttcode("n")` to `r ttcode("n[j]")` in the inner loop to ensure that the `r ttcode("j")`$^{th}$ element of `r ttcode("n")` is used. In the simulation, we use sample sizes of $100, 250, 1000$ and $3000$. Consequently we have a total of four distinct simulations using different sample sizes.

<div class="unfolded"> 
```{r, fig.align='center', cache=T}
# set seed for reproducibility
set.seed(1)

# set repetitions and the vector of sample sizes
reps <- 1000
n <- c(100, 250, 1000, 3000)

# initialize the matrix of outcomes
fit <- matrix(ncol = 2, nrow = reps)

# divide the plot panel in a 2-by-2 array
par(mfrow = c(2, 2))

# loop sampling and plotting

# outer loop over n
for (j in 1:length(n)) {
  
  # inner loop: sampling and estimating of the coefficients
  for (i in 1:reps){
    
    sample <- population[sample(1:N, n[j]), ]
    fit[i, ] <- lm(Y ~ X, data = sample)$coefficients
    
  }
  
  # draw density estimates
  plot(density(fit[ ,2]), xlim=c(2.5, 4.5), 
       col = j, 
       main = paste("n=", n[j]), 
       xlab = bquote(hat(beta)[1]))
  
}
```
</div>

We find that, as $n$ increases, the distribution of $\hat\beta_1$ concentrates around its mean, i.e., its variance decreases. Put differently, the likelihood of observing estimates close to the true value of $\beta_1 = 3.5$ grows as we increase the sample size. The same behavior can be observed if we analyze the distribution of $\hat\beta_0$ instead. 

### Simulation Study 3 {-}

Furthermore, (4.1) reveals that the variance of the OLS estimator for $\beta_1$ decreases as the variance of the $X_i$ increases. In other words, as we increase the amount of information provided by the regressor, that is, increasing $Var(X)$, which is used to estimate $\beta_1$, we become more confident that the estimate is close to the true value (i.e., $Var(\hat\beta_1)$ decreases).<br>
We can visualize this by reproducing Figure 4.6 from the book. To do this, we sample observations $(X_i,Y_i)$, $i=1,\dots,100$ from a bivariate normal distribution with 

$$E(X)=E(Y)=5,$$ 
$$Var(X)=Var(Y)=5$$ 
and 
$$Cov(X,Y)=4.$$

Formally, this is written down as

\begin{align}
  \begin{pmatrix}
    X \\
    Y \\
  \end{pmatrix}
  \overset{i.i.d.}{\sim} & \ \mathcal{N} 
  \left[
    \begin{pmatrix}
      5 \\
      5 \\
    \end{pmatrix}, \ 
    \begin{pmatrix}
      5 & 4 \\
      4 & 5 \\
    \end{pmatrix}
  \right]. \tag{4.3}
\end{align}

To carry out the random sampling, we make use of the function `r ttcode("mvrnorm()")` from the package `r ttcode("MASS")` [@R-MASS] which allows to draw random samples from multivariate normal distributions, see `?mvtnorm`. Next, we use `r ttcode("subset()")` to split the sample into two subsets such that the first set, `r ttcode("set1")`, consists of observations that fulfill the condition $\lvert X - \overline{X} \rvert > 1$ and the second set, `r ttcode("set2")`, includes the remainder of the sample. We then plot both sets and use different colors to distinguish the observations.

<div class="unfolded"> 
```{r, fig.align = 'center', cache = T}
# load the MASS package
library(MASS)

# set seed for reproducibility
set.seed(4)

# simulate bivarite normal data
bvndata <- mvrnorm(100, 
                mu = c(5, 5), 
                Sigma = cbind(c(5, 4), c(4, 5))) 

# assign column names / convert to data.frame
colnames(bvndata) <- c("X", "Y")
bvndata <- as.data.frame(bvndata)

# subset the data
set1 <- subset(bvndata, abs(mean(X) - X) > 1)
set2 <- subset(bvndata, abs(mean(X) - X) <= 1)

# plot both data sets
plot(set1, 
     xlab = "X", 
     ylab = "Y", 
     pch = 19)

points(set2, 
       col = "steelblue", 
       pch = 19)
```
</div>

It is clear that observations that are close to the sample average of the $X_i$ have less variance than those that are farther away. Now, if we were to draw a line as accurately as possible through either of the two sets it is intuitive that choosing the observations indicated by the black dots, i.e., using the set of observations which has larger variance than the blue ones, would result in a more precise line. Now, let us use OLS to estimate slope and intercept for both sets of observations. We then plot the observations along with both regression lines.

<div class="unfolded"> 
```{r, fig.align='center'}
# estimate both regression lines
lm.set1 <- lm(Y ~ X, data = set1)
lm.set2 <- lm(Y ~ X, data = set2)

# plot observations
plot(set1, xlab = "X", ylab = "Y", pch = 19)
points(set2, col = "steelblue", pch = 19)

# add both lines to the plot
abline(lm.set1, col = "green")
abline(lm.set2, col = "red")
```
</div>

Evidently, the green regression line does far better in describing data sampled from the bivariate normal distribution stated in (<a href="#mjx-eqn-4.3">4.3</a>) than the red line. This is a nice example for demonstrating why we are interested in a high variance of the regressor $X$: more variance in the $X_i$ means more information from which the precision of the estimation benefits.

## Exercises

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 1. Class Sizes and Test Scores {-}

A researcher wants to analyze the relationship between class size (measured by the student-teacher ratio) and the average test score. Therefore he measures both variables in $10$ different classes and ends up with the following results.

<!--html_preserve-->

  <table>
      <tr>
        <td><b>Class Size</b></td>
        <td>23</td>
        <td>19</td>
        <td>30</td>
        <td>22</td>
        <td>23</td>
        <td>29</td>
        <td>35</td>
        <td>36</td>
        <td>33</td>
        <td>25</td>
      </tr>
      <tr>
        <td><b>Test Score</b></td>
        <td>430</td>
        <td>430</td>
        <td>333</td>
        <td>410</td>
        <td>390</td>
        <td>377</td>
        <td>325</td>
        <td>310</td>
        <td>328</td>
        <td>375</td>
      </tr>
    </table>

<!--/html_preserve-->

**Instructions:**

+ Create the vectors <tt>cs</tt> (the class size) and <tt>ts</tt> (the test score), containing the observations above.  

+ Draw a scatterplot of the results using <tt>plot()</tt>.

<iframe src="DCL/ex4_1.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>


</div>')
} else {
  cat('\\begin{center}\\textit{This interactive part of the book is only available in the HTML version.}\\end{center}')
}
```

```{r, echo=FALSE, results='asis'}
if (my_output=='html') {
  cat('
<div  class = "DCexercise">

#### 2. Mean, Variance, Covariance and Correlation {-}

The vectors <tt>cs</tt> and <tt>ts</tt> are available in the working environment (you can check this: type their names into the console and press enter).

**Instructions:**

+ Compute the mean, the sample variance and the sample standard deviation of <tt>ts</tt>. 

+ Compute the covariance and the correlation coefficient for <tt>ts</tt> and <tt>cs</tt>.

<iframe src="DCL/ex4_2.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

**Hint:** Use the <tt>R</tt> functions presented in this chapter: <tt>mean()</tt>, <tt>sd()</tt>, <tt>cov()</tt>, <tt>cor()</tt> and <tt>var()</tt>.

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 3. Simple Linear Regression {-}

The vectors <tt>cs</tt> and <tt>ts</tt> are available in the working environment. 

**Instructions:**

+ The function <tt>lm()</tt> is part of the package <tt>AER</tt>. Attach the package using <tt>library()</tt>.

+ Use <tt>lm()</tt> to estimate the regression model $$TestScore_i = \\beta_0 + \\beta_1 STR_i + u_i.$$ Assign the result to <tt>mod</tt>.

+ Obtain a statistical summary of the model.

<iframe src="DCL/ex4_3.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 4. The Model Object {-}

Let us see how an object of class <tt>lm</tt> is structured. 

The vectors <tt>cs</tt> and <tt>ts</tt> as well as the model object <tt>mod</tt> from the previous exercise are available in your workspace.

**Instructions:**

+ Use <tt>class()</tt> to learn about the class of the object <tt>mod</tt>.
+ <tt>mod</tt> is an object of type <tt>list</tt> with named entries. Check this using the function <tt>is.list()</tt>.
+ See what information you can obtain from <tt>mod</tt> using <tt>names()</tt>. 
+ Read out an arbitrary entry of the object <tt>mod</tt> using the <tt>$</tt> operator.

<iframe src="DCL/ex4_4.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 5. Plotting the Regression Line {-}

You are provided with the code for the scatterplot in <tt>script.R</tt>

**Instructions:**

+ Add the regression line to the scatterplot from a few exercises before.

+ The object <tt>mod</tt> is available in your working environment.

<iframe src="DCL/ex4_5.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

**Hint:** Use the function <tt>abline()</tt>.

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 6. Summary of a Model Object {-}

Now read out and store some of the information that is contained in the output of <tt>summary()</tt>.

**Instructions:**

+ Assign the output of <tt>summary(mod)</tt> to the variable <tt>s</tt>.

+ Check entry names of the object <tt>s</tt>.

+ Create a new variable <tt>R2</tt> and assign the $R^2$ of the regression.

The object <tt>mod</tt> is available in your working environment.

<iframe src="DCL/ex4_6.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 7. Estimated Coefficients {-}

The function <tt>summary()</tt> also provides information on the statistical significance of the estimated coefficients.

**Instructions:**

Extract the named $2\\times4$ matrix with estimated coefficients, standard errors, $t$-statistics and corresponding $p$-values from the model summary <tt>s</tt>. Save this matrix in an object named <tt>coefs</tt>.

The objects <tt>mod</tt> and <tt>s</tt> are available in your working environment.

<iframe src="DCL/ex4_7.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 8. Dropping the Intercept {-}

So far, we have estimated regression models consisting of an intercept and a single regressor. In this exercise you will learn how to specify and how to estimate regression a model without intercept.

Note that excluding the intercept from a regression model might be a dodgy practice in some applications as this imposes the conditional expectation function of the dependent variable to be zero if the regressor is zero.

**Instructions:**

+ Figure out how the <tt>formula</tt> argument must be specified for a regression of <tt>ts</tt> solely on <tt>cs</tt>, i.e., a regression without intercept. Google is your friend!

+ Estimate the regression model without intercept and store the result in <tt>mod_ni</tt>.

The vectors <tt>cs</tt>, <tt>ts</tt> and the model object <tt>mod</tt> from previous exercises are available in the working environment.

<iframe src="DCL/ex4_8.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 9. Regression Output: No Constant Case {-}

In Exercise 8 you have estimated a model without intercept. The estimated regression function is

$$\\widehat{TestScore} = \\underset{(1.36)}{12.65} \\times STR.$$

**Instructions:**

Convince yourself that everything is as stated above: extract the coefficient matrix from the summary of <tt>mod_ni</tt> and store it in a variable named <tt>coef</tt>.

The vectors <tt>cs</tt>, <tt>ts</tt> as well as the model object <tt>mod_ni</tt> from the previous exercise are available in your working environment.

<iframe src="DCL/ex4_9.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

**Hint:** An entry of a named list can be accessed using the <tt>$</tt> operator.

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 10. Regression Output: No Constant Case --- Ctd. {-}

In Exercises 8 and 9 you have dealt with a model without intercept. The estimated regression function was

$$\\widehat{TestScore_i} = \\underset{(1.36)}{12.65} \\times STR_i.$$

The coefficient matrix <tt>coef</tt> from Exercise 9 contains the estimated coefficient on $STR$, its standard error, the $t$-statistic of the significance test and the corresponding $p$-value.

**Instructions:**

+ Print the contents of <tt>coef</tt> to the console.
+ Convince yourself that the reported $t$-statistic is correct: use the entries of <tt>coef</tt> to compute the $t$-statistic and save it to <tt>t_stat</tt>.

The matrix <tt>coef</tt> from the previous exercise is available in your working environment.

<iframe src="DCL/ex4_10.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

**Hints:** 

+ <tt>X[a,b]</tt> returns the <tt>[a,b]</tt> element of the matrix <tt>X</tt>. 

+ The $t$-statistic for a test of the hypothesis $H_0: \\beta_1 = 0$ is computed as $$t = \\frac{\\hat{\\beta}_1}{SE(\\hat{\\beta}_1)}.$$

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 11. Two Regressions, One Plot {-}

The two estimated regression models from the previous exercises are

$$\\widehat{TestScore_i} = \\underset{(1.36)}{12.65} \\times STR_i$$

and

$$\\widehat{TestScore_i} = \\underset{(23.96)}{567.4272} \\underset{(0.85)}{-7.1501} \\times STR_i.$$

You are provided with the code line <tt>plot(cs, ts)</tt> which creates a scatterplot of <tt>ts</tt> and <tt>cs</tt>. Note that this line must be executed before calling <tt>abline()</tt>! You may color the regression lines by using, e.g., <tt>col = "red"</tt> or <tt>col = "blue"</tt> as an additional argument to <tt>abline()</tt> for better distinguishability.

The vectors <tt>cs</tt> and <tt>ts</tt> as well as the list objects <tt>mod</tt> and <tt>mod_ni</tt> from previous exercises are availabe in your working environment.

**Instructions:**

Generate a scatterplot of <tt>ts</tt> and <tt>cs</tt> and add the estimated regression lines of <tt>mod</tt> and <tt>mod_ni</tt>.

<iframe src="DCL/ex4_11.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 12. $TSS$ and $SSR$ {-}

If graphical inspection does not help, researchers resort to analytic techniques in order to detect if a model fits the data at hand good or better than another model.

Let us go back to the simple regression model including an intercept. The estimated regression line for <tt>mod</tt> was

$$\\widehat{TestScore_i} = 567.43 - 7.15 \\times STR_i, \\, R^2 = 0.8976, \\, SER=15.19.$$

You can check this as <tt>mod</tt> and the vectors <tt>cs</tt> and <tt>ts</tt> are available in your working environment.

**Instructions:**

+ Compute $SSR$, the sum of squared residuals, and save it to <tt>ssr</tt>.
+ Compute $TSS$, the total sum of squares, and save it to <tt>tss</tt>.

<iframe src="DCL/ex4_12.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 13. The $R^2$ of a Regression Model {-}

The $R^2$ of the regression saved in <tt>mod</tt> is $0.8976$. You can check this by executing <tt>summary(mod)$r.squared</tt> in the console below.

Remember the formula of $R^2$:

$$R^2 = \\frac{ESS}{TSS} = 1 - \\frac{SSR}{TSS}$$

The objects <tt>mod</tt>, <tt>tss</tt> and <tt>ssr</tt> from the previous exercise are available in your working environment.

**Instructions:**

+ Use <tt>ssr</tt> and <tt>tss</tt> to compute $R^2$ manually. *Round* the result to *four* decimal places and save it to <tt>R2</tt>.
+ Use the logical operator <tt>==</tt> to check whether your result matches the value mentioned above.

<iframe src="DCL/ex4_13.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

**Hints:**

You may round numeric values using the function <tt>round()</tt>.

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 14. The Standard Error of The Regression {-}

The standard error of the Regression in the simple regression model is $$SER = \\frac{1}{n-2} \\sum_{i=1}^n \\widehat{u}_i^2 =\\sqrt{\\frac{SSR}{n-2}}.$$ $SER$ measures the size of an average residual which is an estimate of the magnitude of a typical regression error.

The model object <tt>mod</tt> and the vectors <tt>cs</tt> and </tt>ts</tt> are available in your workspace.

**Instructions:**

+ Use <tt>summary()</tt> to obtain the $SER$ for the regression of <tt>ts</tt> on <tt>cs</tt> saved in the model object <tt>mod</tt>. Save the result in the variable <tt>SER</tt>.

+ Use <tt>SER</tt> to compute the $SSR$ and store it in <tt>SSR</tt>.

+ Check that <tt>SSR</tt> is indeed the $SSR$ by comparing <tt>SSR</tt> to the result of <tt>sum(mod$residuals^2)</tt>

<iframe src="DCL/ex4_14.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

</div>')}
```

```{r, echo=FALSE, results='asis'}
if (my_output=="html") {
  cat('
<div  class = "DCexercise">

#### 15. The Estimated Covariance Matrix {-}

As has been discussed in Chapter \\@ref(tlsa), the OLS estimators $\\widehat{\\beta}_0$ and $\\widehat{\\beta}_1$ are functions of the random error term. Therefore, they are random variables themselves. For two or more random variables, their covariances and variances are summarized by a *variance-covariance matrix* (which is often simply called the *covariance matrix*). Taking the square root of the diagonal elements of the estimated covariance matrix obtains $SE(\\widehat\\beta_0)$ and $SE(\\widehat\\beta_1)$, the standard errors of $\\widehat{\\beta}_0$ and $\\widehat{\\beta}_1$. 

<tt>summary()</tt> computes an estimate of this matrix. The respective entry in the output of summary (remember that <tt>summary()</tt> produces a list) is called <tt>cov.unscaled</tt>. The model object <tt>mod</tt> is available in your workspace.

**Instructions:**

+ Use <tt>summary()</tt> to obtain the covariance matrix estimate for the regression of test scores on student-teacher ratios stored in the model object <tt>mod</tt>. Save the result to <tt>cov_matrix</tt>.

+ Obtain the diagonal elements of <tt>cov_matrix</tt>, compute their square root and assign the result to the variable <tt>SEs</tt>.

<iframe src="DCL/ex4_15.html" frameborder="0" scrolling="no" style="width:100%;height:340px"></iframe>

**Hint:** <tt>diag(A)</tt> returns a vector containing the diagonal elements of the matrix <tt>A</tt>.

</div>')}
```

